Binary Latent Representations for Efficient Ranking

### 2017-06-21 paper/paper.tex
Revise and shorten paper.

\title{Binary Latent Representations for Efficient Ranking}
\author{Maciej Kula}
\email{maciej.kula@gmail.com}
\date{\today}
\settopmatter{printacmref=false, printccs=false, printfolios=true}
\setcopyright{none}
\acmConference[]{}{}{}

\begin{abstract}
Large-scale recommender systems often face severe latency and storage constraints at prediction time. These are particularly acute when the number of items that could be recommended is large, and calculating predictions for the full set is computationally intensive. To relax these constraints, we train recommendation models that use binary rather than real-valued user and item representations, and show that they offer a range of attractive speed-accuracy trade-offs. In particular, we demonstrate 20 times faster prediction at the expense of a modest 7\% loss in ranking accuracy on the Movielens 1M dataset.
\end{abstract}

@@

To alleviate these constraints in both online and offline settings, we propose the use of binary item and user representations in learning-to-rank matrix factorization models. Following the approach of \citet{rastegari2016xnor}, we show that we can estimate binary user and item representations that achieve comparable ranking accuracy to standard learning-to-rank approaches while being several times faster to score, requiring a fraction of the memory, and being only marginally more complex to estimate. In MovieLens 1M experiments, our binary representations are within 4\% of the accuracy of the real-valued model, but are over 13 times faster to score and take only 6\% of the memory.

\section{Our approach}
\label{sec:approach}
\citet{rastegari2016xnor} introduce XNOR-Networks, an approach that combines 1-bit quantization of fully-connected and convolutional layers with representing dot products through scaled XNOR and bitcounting operations.

@@

Naturally, using binary instead of real-valued representations entails a loss of representational power, and the resulting models can be expected to be less accurate. Nevertheless, binary models may still be very useful in a production system, for two reasons. Firstly, they may offer the system designer a more favourable trade-off between speed and accuracy than simply changing the number of latent dimensions in a real-valued model. If latency is the binding constraint, switching to a binary model may offer better or comparable speed with better ranking performance. If accuracy is the priority, a binary model may offer equivalent ranking performance while being faster to score. This suspicion is borne out in our experiments. Secondly, many production systems resort to heavy use of heuristics when selecting candidates to be scored by their ranking systems (as scoring all candidates would be infeasible). To the extent that these heuristics are suboptimal, replacing them with a binary representation model would improve the overall accuracy of the system.

@@

\citet{rastegari2016xnor} show that the dot product of two real-valued vectors can be approximated in the binary domain by (1) binarizing the input vectors using the sign function (so that the results lie in $\{1, -1\}$), (2) taking their dot product, and (3) scaling the result by the average magnitude of the vectors' elements. Letting
\begin{equation}
\alpha = \frac{1}{n}\norm{\vec{i}_{i}}_{\ell1}
\end{equation}
and
\begin{equation}
\beta = \frac{1}{n}\norm{\vec{u}_{u}}_{\ell1}
\end{equation}
represent the scaling factors, the approximation is given by
\newcommand\binaryApproximation{ \alpha \beta\left(\sign(\vec{u}_u) \cdot \sign(\vec{i}_i)\right)}
\begin{equation}
\vec{u}_u \cdot \vec{i}_i \approx \binaryApproximation
\end{equation}
Applying this to the latent-factor model, the prediction for a user-item pair is given by
\begin{equation}
r_{ui} = \binaryApproximation + b_u + b_i
\end{equation}
Note that the scaling factors $\alpha$ and $\beta$ need to be stored (in addition to the latent representations and biases) to compute predictions in the binary model.

@@

For the purposes of this paper, we focus on a simple bilinear collaborative filtering model, and do not use any external metadata information or model the sequential nature of the data. However, the binary dot product approach generalizes beyond simple factorization models, and can be applied as a component of any model whose final scoring step involves a dot product between user and item representations. In particular, models using recurrent \citep{hidasi2015session} or convolutional \citep{lynch2015images} item or user representations can easily be augmented to use binary dot products in the final ranking stages. Assuming item representations can be computed in advance, even very complicated models can enjoy the efficiency advantages of using binary representations.

\section{Experiments}
\begin{table*}[htbp]
\begin{threeparttable}
\caption{Movielens 1M results}
\label{tb:results}
\centering
\begin{tabular}{rrrrrrrr}
\toprule
Dimension &   MRR & Binary MRR & MRR ratio\tnote{1} &   PPMS\tnote{2} & Binary PPMS & PPMS ratio & Memory use ratio \\
\midrule
       32 & 0.072 &      0.059 &     0.816 & 85,510 &     264,310 &      3.091 &            0.091 \\
       64 & 0.075 &      0.065 &     0.862 & 44,964 &     221,000 &      4.915 &            0.062 \\
      128 & 0.076 &      0.070 &     0.926 & 22,394 &     166,273 &      7.425 &            0.047 \\
      256 & 0.078 &      0.072 &     0.916 & 12,040 &     189,353 &     15.727 &            0.039 \\
      512 & 0.079 &      0.073 &     0.921 &   5933 &     118,895 &     20.039 &            0.035 \\
     1024 & 0.081 &      0.076 &     0.933 &   2989 &      59,612 &     19.942 &            0.033 \\
     2048 & 0.080 &      0.075 &     0.930 &   1514 &      32,092 &     21.185 &            0.032 \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small{
\item[1] Ratio of binary model MRR to real-valued model MRR
\item[2] Predictions Per Millisecond
}
\end{tablenotes}
\end{threeparttable}
\end{table*}
To assess the accuracy-speed trade-offs enabled by the XNOR approach, we conduct an experiment on the Movielens 1M dataset \citep{harper2016movielens}. The dataset contains 1 million ratings from 6000 users on 4000 movies. Because the computational speed improvements offered by the binary representations are linear in the size of the catalogue, the results on this dataset should be indicative of results that can be obtained on larger datasets. At the same time, the dataset's small size enables us to easily test hyperparameter configurations and estimate models with high latent dimensions.

\subsection{Experimental setup}
We randomly divide the dataset into a training, test, and validation set. In order to build a picture of the accuracy-speed trade-offs afforded by real-valued and binary latent models, we explicitly build models for between 32 and 2048 latent dimensions. For every latent dimensionality, we conduct a random search over the hyperparameter space, and pick optimal learning rates and minibatch sizes for each algorithm based on their performance on the test set. The final results are obtained from ranking interactions from the validation set. We use mean reciprocal rank (MRR) as a measure of ranking quality.

Benchmark results are measured on an Intel Xeon E5-2686v4 CPU by running a 100 repetitions of scoring 100,000 items and averaging the results.

@@

During training, the embedding parameters are stored and updated as single precision floating point values. Similarly, the XNOR dot product is carried out using floats in $\{1, -1\}$. The minibatch size and number of training epochs are treated as model hyperparameters. All models are trained using the Adam training rate schedule \citep{kingma2014adam}.

@@

As expected, for the same dimensionality a binary model achieves lower ranking accuracy. On average, the accuracy loss is around 10\%, varying between 7\% and 18\%.
Nevertheless, the binary model's speed enables some very favourable tradeoffs

When optimizing for MRR in both settings (at 1024 dimensions for the continuous and 1024 for the binary case), the XNOR representation achieves 93\% of the maximum MRR of the float representations at 20 times the query throughput and using only 3\% of the memory.

When optimizing for query throughput at 32 dimensions for the continuous case, the XNOR representation (at 512 dimensions) can improve the ranking performance by achieving 101\% of the continuous MRR while being slightly faster at 1.4 times the query throughput, and using 55\% of the memory.

\section{Related work}
Much of existing work on efficient recommendation retrieval focuses on reducing the number of dot products that need to be performed in order to retrieve top recommendations. \citet{koenigstein2012efficient} introduces an exact branch-and-bound based on metric trees as well as an approximate algorithm that clusters users and serves recommendations computed for the cluster centers. \citet{shrivastava2014asymmetric} extend well-known \citep{indyk1998approximate} locality-sensitive hashing techniques to maximum inner product search (MIPS) through asymmetric transformations of the query and candidate vectors.

Our work complements these approaches. Binary representations are direct replacement for existing real-valued recommendations, and will serve to speed up the final predictions on the set of candidates returned from tree nodes or hash buckets.

The method that is most closely related to the model we present in this paper is \citet{Shen_2015_ICCV}, who expand on MIPS by learning asymmetric binary hash functions. The hash functions are trained to minimize the L2 norm between the inner product matrix of the original input vectors and the inner product matrix of their binary representations. As a post-processing technique, it is a applicable to a wide range of problems. However, because the binary representations are obtained as a separate step, this approach may not be able to exploit the benefits in training a binary representation approach end-to-end.

\section{Conclusion and future work}
We show that binary user and item representations allow recommender system designers to exploit an attractive new set of speed-accuracy trade-offs. They are straightforward to estimate, and can easily be used in any existing model estimated using stochastic gradient descent.

>>>>

\title{Binary Latent Representations for Efficient Ranking}
\subtitle{Empirical Assessment}
\author{Maciej Kula}
\email{maciej.kula@gmail.com}
\date{\today}
\settopmatter{printacmref=false, printccs=false, printfolios=true}
\setcopyright{none}
\acmConference[]{}{}{}

\begin{abstract}
Large-scale recommender systems often face severe latency and storage constraints at prediction time. These are particularly acute when the number of items that could be recommended is large, and calculating predictions for the full set is computationally intensive. In an attempt to relax these constraints, we train recommendation models that use binary rather than real-valued user and item representations, and show that while they are substantially faster to evaluate, the gains in speed come at a large cost in accuracy. In our Movielens 1M experiments, we show that reducing the latent dimensionality of traditional models offers a more attractive accuracy/speed trade-off than using binary representations.
\end{abstract}

@@

To alleviate these constraints in both online and offline settings, we evaluate the use of binary item and user representations in learning-to-rank matrix factorization models. Following the approach of \citet{rastegari2016xnor}, we estimate binary user and item representations that are several times faster to score and requiring a fraction of the memory (for the same latent dimensionality). Unfortunately, they are substantially less accurate than standard learning-to-rank approaches. The resulting speed-accuracy trade-offs favour simply reducing the latent dimensionality of traditional models when evaluation speed is desired.

We speculate that binary representations are more useful in case of more complex models (such as deep convolutional models) where a single prediction requires many more floating operations than relatively compact biliearn recommender models.

%%_______________________________________
%% In MovieLens 1M experiments, our binary representations are within 4\% of the accuracy of the real-valued model, but are over 13 times faster to score and take only 6\% of the memory.
%% ____________________________________

\section{Binary latent representations}
\label{sec:approach}
\citet{rastegari2016xnor} introduce XNOR-Networks, an approach that combines 1-bit quantization of fully-connected and convolutional layers with representing dot products through scaled XNOR and bitcounting operations.

@@

Naturally, using binary instead of real-valued representations entails a loss of representational power, and the resulting models can be expected to be less accurate. Nevertheless, binary models may still be very useful in a production system, for two reasons. Firstly, they may offer the system designer a more favourable trade-off between speed and accuracy than simply changing the number of latent dimensions in a real-valued model. If latency is the binding constraint, switching to a binary model may offer better or comparable speed with better ranking performance. If accuracy is the priority, a binary model may offer equivalent ranking performance while being faster to score. Secondly, many production systems resort to heavy use of heuristics when selecting candidates to be scored by their ranking systems (as scoring all candidates would be infeasible). To the extent that these heuristics are suboptimal, replacing them with a binary representation model would improve the overall accuracy of the system.

@@

\citet{rastegari2016xnor} show that the dot product of two real-valued vectors can be approximated in the binary domain by (1) binarizing the input vectors using the sign function (so that the results lie in $\{1, -1\}$), (2) taking their dot product, and (3) scaling the result by the average magnitude of the vectors' elements. Letting
\begin{equation}
\alpha = \frac{1}{n}\norm{\vec{i}_{i}}_{\ell1}
\end{equation}
and
\begin{equation}
\beta = \frac{1}{n}\norm{\vec{u}_{u}}_{\ell1}
\end{equation}
represent the scaling factors, the approximation is given by
\newcommand\binaryApproximation{ \alpha \beta\left(\sign(\vec{u}_u) \cdot \sign(\vec{i}_i)\right)}
\begin{equation}
\vec{u}_u \cdot \vec{i}_i \approx \binaryApproximation
\end{equation}
Applying this to the latent-factor model, the prediction for a user-item pair is given by
\begin{equation}
r_{ui} = \binaryApproximation + b_u + b_i
\end{equation}
Note that the scaling factors $\alpha$ and $\beta$ need to be stored (in addition to the latent representations and biases) to compute predictions in the binary model. Their use also implies a constant cost of two floating-point multiplications per dot product when using binary representations.

@@

For the purposes of this paper, we focus on a simple bilinear collaborative filtering model, and do not use any external metadata information or model the sequential nature of the data. However, the binary dot product approach generalizes beyond simple factorization models, and can be applied as a component of any model whose final scoring step involves a dot product between user and item representations. In particular, models using recurrent \citep{hidasi2015session} or convolutional \citep{lynch2015images} item or user representations can easily be augmented to use binary dot products in the final ranking stages.

\section{Experiments}
\begin{table*}[htbp]
\begin{threeparttable}
\caption{Movielens 1M results}
\label{tb:results}
\centering
\begin{tabular}{rrrrrrrr}
\toprule
Dimension &   MRR & Binary MRR & MRR ratio\tnote{1} &   PPMS\tnote{2} & Binary PPMS & PPMS ratio & Memory use ratio \\
\midrule
       32 & 0.077 &      0.059 &     0.768 & 87,758 &     264,146 &      3.010 &            0.091 \\
       64 & 0.075 &      0.066 &     0.878 & 45,992 &     220,833 &      4.802 &            0.062 \\
      128 & 0.076 &      0.071 &     0.935 & 23,870 &     166,252 &      6.965 &            0.047 \\
      256 & 0.078 &      0.071 &     0.908 & 12,284 &     190,131 &     15.477 &            0.039 \\
      512 & 0.080 &      0.073 &     0.912 &   6074 &     122,637 &     20.190 &            0.035 \\
     1024 & 0.080 &      0.075 &     0.936 &   3056 &      61,301 &     20.056 &            0.033 \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small{
\item[1] Ratio of binary model MRR to real-valued model MRR
\item[2] Predictions Per Millisecond
}
\end{tablenotes}
\end{threeparttable}
\end{table*}
To assess the accuracy-speed trade-offs enabled by the XNOR approach, we conduct an experiment on the Movielens 1M dataset \citep{harper2016movielens}. The dataset contains 1 million ratings from 6000 users on 4000 movies. Because the computational speed improvements offered by the binary representations are linear in the size of the catalogue, the results on this dataset should be indicative of results that can be obtained on larger datasets. At the same time, the dataset's small size enables us to easily test hyperparameter configurations and estimate models with high latent dimensions.

\subsection{Experimental setup}
We randomly divide the dataset into a training, test, and validation set. In order to build a picture of the accuracy-speed trade-offs afforded by real-valued and binary latent models, we explicitly build models for between 32 and 1024 latent dimensions. For every latent dimensionality, we conduct a random search over the hyperparameter space, and pick optimal initial learning rate, loss function, L2 penalty, minibatch size and number of training epochs for each algorithm based on their performance on the test set. The final results are obtained from ranking interactions from the validation set. We use mean reciprocal rank (MRR) as a measure of ranking quality.

Benchmark results are measured on an Intel Xeon E5-2686v4 CPU by running a 500 repetitions of scoring 100,000 items and averaging the results.

@@

During training, the embedding parameters are stored and updated as single precision floating point values. Similarly, the XNOR dot product is carried out using floats in $\{1, -1\}$. The initial learning rate, loss function, L2 penalty, minibatch size and number of training epochs are treated as model hyperparameters. All models are trained using the Adam training rate schedule \citep{kingma2014adam}.

@@

As expected, for the same dimensionality a binary model achieves lower ranking accuracy. On average, the accuracy loss when moving from a contious to a binary model of the same latent dimensionality is around 11\%, varying between 6\% and 23\%.

Unfortunately, while continous models retain good accuracy as latent dimensionality decreases, binary models' representational power sharply deteriorates. Moving from the 1024 to 32 dimensions in the continuous model implies a 29 times increase in prediction speed at the expense of a modest 4\% decrease in accuracy. This compares favourably to switching to binary representations: moving to a 1024-binary dimensional representation implies a sharper accuracy drop at 6\% in exchange for a smaller 20 times increase in prediction speed. Moving to 32 dimensions yields a further speed gains at 86 times, but at the cost of a considerable loss of accuracy at 26\%.

The results suggest that latent recommender models have good representational power even at low latent dimensionalities. This advantage is lost when using binary embeddings, which need to be high-dimensional to achieve comparable accuracy. At the same time, binary representations' speed advantage is only evident at high dimensions: at low dimensions, only floating-point operations can use enjoy the advantages of SIMD operations, and the fixed overhead of binary scaling factors $\alpha$ and $\beta$ constitutes a larger proportion of the total computational cost. We conjecture that the attractiveness of binary representations is tightly coupled with high-dimensional models where a single prediction requires many floating point operations, such as convolutional neural networks. Relatively compact latent factor models do not fall into this category.

\section{Conclusion}
While prediction latency is a pressing concern for many recommender systems, we find that the already compact latent factor models do not stand to benefit from using binary latent representations. The sharp drop in representational power exhibited by binary embeddings makes other approaches (such as exact \citep{koenigstein2012efficient} and approximate \citet{shrivastava2014asymmetric} maximum inner product search) more promising avenues when optimizing large-scale ranking.
